---
title: "Getting Started with networkR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with networkR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width  = 8,
  fig.height = 5.5,
  out.width  = "100%"
)
library(networkR)
```

## What this vignette does

This vignette walks you through every stage of the `networkR` pipeline using
**fully synthetic data** that mimics a multi-tissue malaria vaccine immunology study.
No real data files are needed — the mock panels are generated inline and written to a
temporary directory.  By the end you will have:

- generated and inspected raw long-format flow-cytometry panels;
- configured a pipeline for two tissue compartments and two stratifications;
- traced data through imputation, discretization, and blacklist construction;
- learned and plotted Bayesian network structure on the full cohort;
- compared network structure between exposure groups (Malaria Preexposed vs. Naive) 
  and between protection outcomes (Protected vs. Not protected).

---

## 1. Background: Bayesian Networks for immunology

### 1.1 What is a Bayesian Network?

A **Bayesian Network (BN)** is a probabilistic graphical model whose structure is a
directed acyclic graph (DAG).  Each node $X_i$ represents a variable; each directed
edge $X_j \to X_i$ encodes that $X_j$ is a direct probabilistic predictor of $X_i$
after conditioning on all other parents.  The joint distribution factorises along
the graph:

$$P(X_1, \dots, X_n) = \prod_{i=1}^{n} P\!\left(X_i \mid \text{pa}(X_i)\right)$$

where $\text{pa}(X_i)$ is the **parent set** of $X_i$.  Missing edges denote
*conditional independence*: $X_i \perp\!\!\!\perp X_j \mid \{X_k : k \ne i, j\}$.  This
makes BNs useful for teasing apart direct from indirect associations in high-dimensional
immunological data.

### 1.2 Variable tiers

`networkR` formalises a **tier system** that encodes causal precedence between
variable classes.  Arcs that would run from a lower-priority tier into a
higher-priority tier are *blacklisted* before structure learning begins.

| Tier | Variable class | Examples | Rationale |
|------|----------------|---------|-----------|
| 1 | Exogenous / design | `GroupName`, `VaccinationStatus`, `Timepoint` | Fixed before any biological measurement |
| 2 | Upstream tissue measurements | `Liver_Pct_*` | Liver is exposed to malaria parasites first |
| 3 | Downstream tissue measurements | `PBMC_Pct_*`, `BM_Pct_*` | Systemic immunity reflects hepatic priming |

Three blacklist components are constructed automatically:

1. **Measurement → Exogenous** — no cellular proportion can cause a design variable.
2. **Exogenous → Exogenous** — design factors are mutually independent by construction
   (factorial design assumption; configurable).
3. **Cross-tier (high tier → low tier)** — e.g. PBMC cells (tier 3) cannot be parents
   of Liver cells (tier 2).

Compartment identity is encoded structurally through the measurement column prefixes
and `measurementGroups` definitions, not as a separate row-level `Tissue` node.
That lets the network learn within- and cross-compartment relationships from
paired subject-level measurements while still using measurement groups for node
colouring and tier-aware blacklist construction.

### 1.3 Bootstrap stability and the averaged network

Because the data are finite, a single structure-learning run is sensitive to sampling
noise.  `networkR` uses **bootstrap resampling** (Friedman et al. 1999): $R$ bootstrap
datasets are drawn and a network is learned on each.  The **arc stability** of an edge
$A \to B$ is the fraction of bootstrap networks that contain $A \to B$ *or* $B \to A$
(direction is assessed separately by the `direction` field of `boot.strength()`).

The **averaged network** retains only arcs whose stability exceeds a threshold
(default 0.40 for real studies; we use 0.20 here for the small demo dataset).

### 1.4 Signed arc strength

Raw bootstrap strength is unsigned.  `networkR` multiplies it by the sign of the
Pearson correlation between the two nodes on continuous (pre-discretization) data,
giving a **signed strength** $s$:

$$s_{AB} = \text{sign}(r_{AB}) \times \hat{p}_{AB}$$

where $\hat{p}_{AB}$ is bootstrap arc probability and $r_{AB}$ is Pearson $r$.
The **delta** $\Delta = s_{\text{group}_1} - s_{\text{group}_2}$ summarises
directional enrichment of an edge between two cohorts.

---

## 2. Mock data: a simulated malaria vaccine study

### 2.1 Study design

The simulated cohort has **24 subjects** divided into three experimental groups, each
measured in two tissue compartments (Liver and PBMC) at a single timepoint (`D1`).
This yields **24 rows** in the joined subject table after `ReadPanels()` — one row
per subject, with Liver and PBMC features occupying separate column groups.

```{r study-design-table, echo=FALSE}
design_tbl <- data.frame(
  Group              = c("Malaria Preexposed", "Malaria Naive", "Control Animal"),
  n                  = c(8, 8, 8),
  `Vaccination status` = c("Vaccinated", "Vaccinated", "Unvaccinated"),
  `Protection outcome` = c("Yes / No (4 each)", "Yes / No (4 each)", "N/A"),
  check.names = FALSE
)
knitr::kable(design_tbl, caption = "Simulated study design (24 subjects total)")
```

Subjects S01–S08 are Malaria Preexposed (vaccinated); S09–S16 are Malaria Naive
(vaccinated); S17–S24 are Control Animals (unvaccinated, no protection outcome).

### 2.2 Data-generation helpers

We simulate cell-type proportions using the **Dirichlet distribution**.  The
group-specific concentration parameters $\boldsymbol{\alpha}$ encode biological
signal: Preexposed animals have relatively elevated Kupffer cells (Liver) and CD8
T cells (PBMC), while Control Animals have flatter distributions consistent with
a naive systemic immune state.

```{r helpers}
# Dirichlet draw for a single observation.
# Exploits: if G_i ~ Gamma(alpha_i, 1) then G / sum(G) ~ Dirichlet(alpha).
rdirichlet_one <- function(alpha) {
  raw <- vapply(alpha, function(a) stats::rgamma(1, shape = a, rate = 1), numeric(1))
  raw / sum(raw)
}

# Build a long-format panel table (one row per subject × cell type).
make_panel_long <- function(cell_types, alpha_by_group,
                             subjects, groups, protection, seed) {
  set.seed(seed)
  rows <- lapply(seq_along(subjects), function(i) {
    fracs <- rdirichlet_one(alpha_by_group[[groups[i]]])
    data.frame(
      SubjectId       = subjects[i],
      TimepointLabel2 = "D1",
      GroupName       = groups[i],
      Protection      = protection[i],
      CellType        = cell_types,
      Fraction        = fracs,
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}
```

### 2.3 Generating the Liver panel

Five liver cell types are simulated.  The Dirichlet $\boldsymbol{\alpha}$ shapes
differ by group: Preexposed animals have a higher $\alpha$ for Kupffer Cells
(resident macrophages that capture malaria parasites), while Naive and
Control animals show more uniform distributions.

```{r liver-panel}
# Subject metadata
n_subjects      <- 24
subject_ids     <- sprintf("S%02d", 1:n_subjects)
group_labels    <- rep(c("Malaria Preexposed", "Malaria Naive", "Control Animal"), each = 8)
protection_raw  <- c(
  rep(c("Yes", "No"), each = 4),  # S01-S08 Preexposed: 4 protected, 4 not
  rep(c("Yes", "No"), each = 4),  # S09-S16 Naive: 4 protected, 4 not
  rep("No", 8)                    # S17-S24 Controls: placeholder → overwritten to NA
)

# Liver cell types and group-specific Dirichlet alpha vectors
liver_cell_types <- c(
  "Kupffer Cells", "CD8 T Cells", "NK Cells",
  "Gamma Delta T Cells", "Conventional DC"
)
liver_alpha <- list(
  "Malaria Preexposed" = c(5.0, 2.0, 2.0, 1.0, 0.8),  # elevated Kupffer + moderate NK
  "Malaria Naive"      = c(2.0, 3.0, 3.0, 1.5, 0.5),  # elevated CD8 / NK
  "Control Animal"     = c(1.0, 1.0, 2.5, 1.0, 0.5)   # relatively flat; more NK
)

liver_panel <- make_panel_long(
  liver_cell_types, liver_alpha, subject_ids, group_labels, protection_raw, seed = 42
)
# Add ~10 % random missing values to simulate instrument dropouts
set.seed(7)
liver_panel$Fraction[sample(nrow(liver_panel), floor(nrow(liver_panel) * 0.10))] <- NA

cat("Liver panel: ", nrow(liver_panel), "rows × ", ncol(liver_panel), "columns\n")
cat("Missing Fraction values:", sum(is.na(liver_panel$Fraction)), "\n")
```

```{r liver-head, echo=FALSE}
knitr::kable(
  head(liver_panel[order(liver_panel$SubjectId, liver_panel$CellType), ], 10),
  digits = 3,
  caption = "First 10 rows of the raw Liver panel (long format)"
)
```

### 2.4 Generating the PBMC panel

```{r pbmc-panel}
pbmc_cell_types <- c(
  "CD14 Monocytes", "CD4 T Cells", "CD8 T Cells", "NK Cells", "B Cells"
)
pbmc_alpha <- list(
  "Malaria Preexposed" = c(2.0, 2.0, 4.5, 2.0, 0.5),  # strong CD8 enrichment
  "Malaria Naive"      = c(3.5, 3.0, 2.0, 2.0, 0.5),  # monocyte / CD4 skew
  "Control Animal"     = c(4.5, 2.0, 1.0, 1.0, 0.5)   # monocyte dominated
)

pbmc_panel <- make_panel_long(
  pbmc_cell_types, pbmc_alpha, subject_ids, group_labels, protection_raw, seed = 99
)
set.seed(8)
pbmc_panel$Fraction[sample(nrow(pbmc_panel), floor(nrow(pbmc_panel) * 0.10))] <- NA

cat("PBMC panel:", nrow(pbmc_panel), "rows × ", ncol(pbmc_panel), "columns\n")
cat("Missing Fraction values:", sum(is.na(pbmc_panel$Fraction)), "\n")
```

### 2.5 Writing panels to disk

`ReadPanels()` discovers files by globbing `dataPath` for filenames matching
`filePattern`.  The tissue compartment is parsed from the filename: any file containing
`"liver"` (case-insensitive) maps to the canonical label `"Liver"`.

```{r write-panels}
panel_dir <- file.path(tempdir(), "networkR_vignette_panels")
dir.create(panel_dir, recursive = TRUE, showWarnings = FALSE)
utils::write.csv(liver_panel, file.path(panel_dir, "liver_panel.csv"), row.names = FALSE)
utils::write.csv(pbmc_panel,  file.path(panel_dir, "pbmc_panel.csv"),  row.names = FALSE)
cat("Panel files written to:", panel_dir, "\n")
list.files(panel_dir)
```

---

## 3. Configuration

### 3.1 Configuration concepts

`networkR` is configured entirely through a named R list produced by
`BuildConfiguration()` (or equivalently `ParseConfiguration()` for a YAML file).
The configuration controls every stage of the pipeline:

| Section | Key parameters |
|---------|---------------|
| `study` | `studyName`, `randomSeed` |
| `data` | `dataPath`, `filePattern`, column name mappings, tissue map |
| `variables` | `exogenousVariables`, `stratifierVariables`, `derivedVariables`, `measurementGroups` |
| `imputation` | `method` (pmm), `numberOfImputations`, `maximumIterations` |
| `discretization` | `method` (hartemink), `initialBreaks`, `finalBreaks`, `fallbackBreaks` |
| `structureLearning` | `algorithms`, `bootstrapReplicates`, `averageNetworkThreshold` |
| `stratifications` | List of cohort stratification definitions |
| `plotting` | Layout, colours, sizes |
| `output` | Save paths and file format |

### 3.2 Building the vignette configuration

We override only the parameters that differ from the package defaults.  Key demo
choices: **50 bootstrap replicates** and **`averageNetworkThreshold = 0.2`** to ensure
a non-trivial averaged network from a small dataset.  A real study should use
$R \ge 1000$ replicates and `averageNetworkThreshold = 0.4`.

```{r configuration}
configuration <- BuildConfiguration(list(
  study = list(
    studyName  = "Malaria Vaccine Immunology — Demo Study",
    randomSeed = 42
  ),
  data = list(
    dataPath    = panel_dir,
    filePattern = "[.](csv)$"
  ),
  variables = list(
    measurementGroups = list(
      list(
        groupName      = "liverCells",
        columnPrefixes = list("Liver_"),
        tier           = 2,
        fillColour     = "lightgreen",
        borderColour   = "darkgreen"
      ),
      list(
        groupName      = "pbmcCells",
        columnPrefixes = list("PBMC_"),
        tier           = 3,
        fillColour     = "lightblue",
        borderColour   = "steelblue"
      )
    )
  ),
  imputation = list(
    numberOfImputations = 3,
    maximumIterations   = 5
  ),
  structureLearning = list(
    algorithms                 = list("tabu"),
    selectedBootstrapAlgorithm = "tabu",
    bootstrapReplicates        = 50,
    averageNetworkThreshold    = 0.2
  ),
  stratifications = list(
    list(
      analysisName = "exposure",
      variableName = "GroupName",
      levels       = list("Malaria Preexposed", "Malaria Naive"),
      filters      = list()
    ),
    list(
      analysisName = "protection",
      variableName = "Protection",
      levels       = list("Yes", "No"),
      filters      = list(
        list(column = "VaccinationStatus", operator = "equals",    value = "Vaccinated"),
        list(column = "Protection",        operator = "notMissing")
      )
    )
  ),
  output = list(savePlots = FALSE)
))

print(configuration)
```

A YAML file equivalent to the above is available at `system.file("configuration_template.yml", package = "networkR")`.

---

## 4. Running the full pipeline

`RunPipeline()` is the one-call entry point.  It executes the complete chain —
ingestion, derived-variable annotation, imputation, discretization, full-model
structure learning, stratified sub-models, and pairwise structure comparisons —
and returns a single `networkRPipelineResult` object containing every intermediate
result and plot.

```{r run-pipeline}
set.seed(42)
pipeline_result <- RunPipeline(configuration)
print(pipeline_result)
```

The remainder of this vignette unpacks each component of `pipeline_result` in detail.

---

## 5. Panel ingestion

### 5.1 How `ReadPanels()` works

`ReadPanels()` discovers all files in `configuration$data$dataPath` whose names
match `configuration$data$filePattern`.  For each file it:

1. Reads the raw long-format table (xlsx or csv).
2. Parses the tissue compartment from the filename using `configuration$data$tissueMap`.
3. Renames `TimepointLabel2 → Timepoint` (or whatever your `rawTimepointColumn`/`analysisTimepointColumn` mapping specifies).
4. Normalises `CellType` strings to valid column names and prefixes them with
   `{Tissue}_{cellTypePrefix}` (e.g. `"Kupffer Cells"` → `"Liver_Pct_Kupffer_Cells"`).
5. Pivots from long to wide, producing one subject × timepoint row per tissue panel.

The individual pivoted tables are then **`full_join`-ed** on the shared key columns
(`SubjectId`, `Timepoint`, `Protection`, `GroupName`).  Because tissue identity
already lives in the measurement column prefixes, the join produces **one row per
subject** with both `Liver_` and `PBMC_` columns aligned side by side.  Missing values
after the join therefore reflect genuine assay dropouts rather than a structural
cross-tissue design artifact.

```{r ingestion-api, eval=FALSE}
# The individual functions — equivalent to what RunPipeline calls internally:
subject_table <- ReadPanels(configuration)
subject_table <- AddDerivedVariables(subject_table, configuration)
```

### 5.2 The joined subject table

```{r show-ingestion}
st <- pipeline_result$subjectTable
cat("Subject table dimensions:", nrow(st), "rows ×", ncol(st), "columns\n\n")

# Column inventory
meas_cols <- grep("^(Liver|PBMC)_", colnames(st), value = TRUE)
cat("Measurement columns (", length(meas_cols), "):\n", sep = "")
cat(paste(meas_cols, collapse = ", "), "\n\n")

# Measurement missingness pattern
na_counts <- colSums(is.na(st[, meas_cols]))
cat("NA counts per measurement column:\n")
print(na_counts)
```

```{r ingestion-head, echo=FALSE}
display_cols <- c("SubjectId", "GroupName", "Protection",
                  "VaccinationStatus", "Timepoint")
knitr::kable(
  head(st[, display_cols], 12),
  caption = "First 12 rows of the joined subject table (design columns only)"
)
```

Each row now represents one subject-timepoint observation with both tissue
compartments present as separate feature blocks.  In this synthetic example, any
remaining `NA` values in the measurement columns arise from the simulated assay
dropouts we injected into the raw panels.

### 5.3 Derived variables

`AddDerivedVariables()` applies two rules defined in `configuration$variables$derivedVariables`:

- **`VaccinationStatus`** (`ifEquals`): set to `"Unvaccinated"` when `GroupName == "Control Animal"`, otherwise `"Vaccinated"`.
- **`Protection`** (`replaceTargetWhenEquals`): overwrite `Protection` with `NA` when
  `GroupName == "Control Animal"`.  The raw CSV value for controls is irrelevant.

```{r derived-vars}
cat("VaccinationStatus distribution:\n")
print(table(st$VaccinationStatus, useNA = "ifany"))
cat("\nProtection distribution:\n")
print(table(st$Protection, useNA = "ifany"))
```

---

## 6. Missing-data imputation

### 6.1 MICE and predictive mean matching

All missing values in the measurement columns are imputed before structure learning.
`networkR` uses **Multiple Imputation by Chained Equations (MICE)** with
**predictive mean matching** (`method = "pmm"` by default).  PMM preserves the
empirical marginal distribution of each variable: the imputed value is always a
real observed value from a donor matched on predicted values.

Because the joined analysis table is now **subject-level**, all compartments for a
given subject sit in the same row.  MICE therefore uses real paired cross-compartment
measurements as predictors when filling assay dropouts, rather than reconstructing an
entire missing tissue block from a row-level `Tissue` indicator.

If MICE fails (e.g. too few rows relative to predictors in a small sub-cohort),
`ImputeMeasurements()` automatically falls back to **column-wise median imputation**
and emits a warning.

```{r imputation-api, eval=FALSE}
# Equivalent single-step call:
imputation_result <- ImputeMeasurements(pipeline_result$subjectTable, configuration)
```

### 6.2 Imputation result

```{r show-imputation}
ir <- pipeline_result$imputationResult

cat("Measurement columns imputed (", length(ir$measurementColumns), "):\n", sep = "")
cat(paste(ir$measurementColumns, collapse = "\n"), "\n\n")

# NA counts before vs. after
n_na_before <- sum(is.na(pipeline_result$subjectTable[, ir$measurementColumns]))
n_na_after  <- sum(is.na(ir$continuousAnalysisData[, ir$measurementColumns]))
cat("NAs before imputation:", n_na_before, "\n")
cat("NAs after  imputation:", n_na_after,  "\n\n")

cat("Active exogenous variables:", paste(ir$exogenousVariables, collapse = ", "), "\n")
cat("Stratifier variables (not imputed):", paste(ir$stratifierVariables, collapse = ", "), "\n")
```

The `continuousAnalysisData` slot holds the imputed continuous values (used later
for computing signed arc strengths via Pearson correlations).  The `imputedData`
slot holds the same values plus the subject-level metadata columns.

---

## 7. Discretization

### 7.1 Hartemink's algorithm

Bayesian network score functions for categorical variables require discrete (factor)
inputs.  `networkR` uses **Hartemink's pairwise mutual-information algorithm**
(`bnlearn::discretize(method = "hartemink")`) to choose cut points that maximise
the pairwise mutual information across all pairs of measurement variables jointly.
This is preferable to per-column quantile binning because it is sensitive to the
covariance structure of the data.

The algorithm requires sufficient within-cell counts for an initial $q$-quantile
preprocessing pass (`initialBreaks = 20` by default).  Columns that cannot satisfy
this requirement (too few distinct values or too few rows) are routed to a
**quantile-based fallback** (`fallbackBreaks = 2`), which simply splits at the median.

```{r discretization-api, eval=FALSE}
# Equivalent:
discretization_result <- DiscretizeMeasurements(imputation_result, configuration)
```

### 7.2 Column routing

```{r show-discretization}
dr <- pipeline_result$discretizationResult

cat("Total measurement columns:", length(dr$measurementColumns), "\n")
cat("Hartemink columns (", length(dr$harteminkColumns), "):\n", sep = "")
cat(paste(dr$harteminkColumns, collapse = "\n"), "\n\n")

if (length(dr$fallbackColumns) > 0) {
  cat("Fallback columns (", length(dr$fallbackColumns), "):\n", sep = "")
  cat(paste(dr$fallbackColumns, collapse = "\n"), "\n")
} else {
  cat("Fallback columns: none — all columns passed the Hartemink pre-check.\n")
}
```

### 7.3 Discrete factor levels

After discretization every measurement column is an ordered factor with
`finalBreaks = 3` levels by default.  The Hartemink algorithm determines
cut points jointly; levels are labelled `1`, `2`, `3` (low to high within each
column's marginal distribution).

```{r show-factor-levels}
first_three <- dr$measurementColumns[1:3]
lvl_display <- lapply(first_three, function(cn) {
  x <- dr$discreteData[[cn]]
  data.frame(
    variable = cn,
    `n levels` = nlevels(x),
    levels = paste(levels(x), collapse = " / "),
    check.names = FALSE
  )
})
knitr::kable(do.call(rbind, lvl_display), caption = "Factor levels for first three measurement columns")
```

```{r show-discrete-head, echo=FALSE}
knitr::kable(
  head(dr$discreteData[, c("GroupName", first_three)], 8),
  caption = "First 8 rows of discrete data (design + first three measurement columns)"
)
```

---

## 8. Blacklist construction

### 8.1 The three blacklist components

`BuildBlacklist()` assembles a `data.frame` of forbidden arcs (`from`, `to`) from
three sources:

1. **Measurement → Exogenous**: Every arc from any measurement column into any active
   exogenous variable is forbidden (cellular proportions cannot *cause* group assignments).
2. **Exogenous → Exogenous**: Arcs between exogenous variables are forbidden (factorial
   design assumption — modify if your exogenous variables are correlated).
3. **Tier cross-arcs**: Arcs from measurement group $A$ into measurement group $B$ are
   forbidden whenever $\text{tier}(A) > \text{tier}(B)$ — i.e. downstream-tissue cells
   (PBMC, tier 3) cannot be parents of upstream-tissue cells (Liver, tier 2).

```{r blacklist-api, eval=FALSE}
# Equivalent:
blacklist_result <- BuildBlacklist(discretization_result, configuration)
```

### 8.2 Examining the blacklist

```{r show-blacklist}
bl <- pipeline_result$fullModelResult$blacklist

cat("Total blacklisted arcs:", nrow(bl), "\n\n")

# Breakdown by arc category
is_exo_src <- bl$from %in% pipeline_result$fullModelResult$activeExogenousVariables
is_exo_tgt <- bl$to   %in% pipeline_result$fullModelResult$activeExogenousVariables

cat("Arcs blocked: measurement → exogenous: ",
    sum(!is_exo_src &  is_exo_tgt), "\n",
    "Arcs blocked: exogenous → exogenous:  ",
    sum( is_exo_src &  is_exo_tgt), "\n",
    "Arcs blocked: high-tier → low-tier:   ",
    sum(!is_exo_src & !is_exo_tgt), "\n", sep = "")
```

```{r blacklist-head, echo=FALSE}
knitr::kable(
  head(bl, 15),
  caption = "First 15 rows of the blacklist (from → to arcs that cannot exist)"
)
```

---

## 9. Structure learning

### 9.1 Tabu search

`networkR` supports Hill-Climbing (`"hillClimbing"`) and **Tabu search** (`"tabu"`)
for score-based BN structure learning.  Tabu search augments greedy hill-climbing
with a recency memory (the *tabu list*) that forbids recently reversed moves,
helping it escape local optima.  Both algorithms optimise the **BIC score** for
discrete data by default.

After learning the single best-scoring network on the full dataset, `networkR`
runs **bootstrap resampling**: it draws $R$ datasets with replacement and re-runs
Tabu search on each, then aggregates arc frequencies into a `boot.strength` table.

```{r structure-api, eval=FALSE}
# Equivalent:
structure_result <- LearnBayesianNetwork(discretization_result, configuration)
```

### 9.2 BIC score

```{r show-bic}
sr <- pipeline_result$fullModelResult
cat("BIC scores by algorithm:\n")
print(round(sr$networkScores, 2))
```

### 9.3 Bootstrap arc strength

Each row in the `bootstrapStrength` table is a directed arc (`from → to`) and its
estimated stability (`strength`) and direction probability (`direction`).  We sort
by stability descending to see the most robustly supported edges.

```{r show-bootstrap}
bs <- as.data.frame(sr$bootstrapStrength)
bs <- bs[order(-bs$strength), ]
knitr::kable(
  head(bs, 15),
  digits = 3,
  row.names = FALSE,
  caption = "Top 15 arcs by bootstrap stability (strength = fraction of bootstrap networks containing the arc)"
)
```

### 9.4 The averaged network

The averaged network retains only arcs whose stability exceeds
`averageNetworkThreshold` (here 0.20).  This is the network visualized and compared
downstream.

```{r show-averaged-network}
cat("Averaged network threshold:", sr$threshold, "\n\n")
avg_arcs <- bnlearn::arcs(sr$averagedNetwork)
cat("Arcs in the averaged network (", nrow(avg_arcs), "):\n", sep = "")
if (nrow(avg_arcs) > 0) {
  knitr::kable(avg_arcs, row.names = FALSE)
} else {
  cat("No arcs exceeded the threshold — try lowering averageNetworkThreshold.\n")
}
```

---

## 10. Network visualization

`PlotBayesianNetwork()` renders the averaged network using `ggraph` with a
Fruchterman-Reingold (`"fr"`) force-directed layout.  Node colours encode
membership in a measurement group or exogenous category; edge width and
transparency encode bootstrap stability.

**Node colour scheme for this demo:**

| Node type | Fill | Border |
|-----------|------|--------|
| Exogenous variables | mistyrose | firebrick |
| Liver cells (`liverCells`) | lightgreen | darkgreen |
| PBMC cells (`pbmcCells`) | lightblue | steelblue |

```{r plot-full-network, fig.width=9.5, fig.height=7}
pipeline_result$plotObjects$fullModel
```

Wider edges and more opaque edges indicate higher bootstrap stability.  The node
layout is stochastic (force-directed); run with `set.seed()` before
`PlotBayesianNetwork()` for a reproducible layout.

---

## 11. Stratified analysis

### 11.1 Exposure stratification: Malaria Preexposed vs. Naive

`RunPipeline()` runs a separate BN on each stratum defined in
`configuration$stratifications`.  For the `"exposure"` stratification, rows are
partitioned by `GroupName` into `"Malaria Preexposed"` and `"Malaria Naive"`
sub-cohorts (8 subjects = 8 rows each).

```{r plot-preexposed, fig.width=9.5, fig.height=7}
pipeline_result$plotObjects$stratifiedModels$exposure$levelPlots[["Malaria Preexposed"]]
```

```{r plot-naive, fig.width=9.5, fig.height=7}
pipeline_result$plotObjects$stratifiedModels$exposure$levelPlots[["Malaria Naive"]]
```

### 11.2 Exposure comparison plot

The comparison scatter plot places **signed bootstrap strength** for Group 1
(Malaria Preexposed) on the y-axis and Group 2 (Malaria Naive) on the x-axis.
Each point is one edge from the union of both groups' bootstrap tables.

- **On the identity line** ($y = x$): equally strong in both groups.
- **Above the identity line**: stronger (or in a different direction) in Preexposed.
- **Below the identity line**: stronger in Naive.

Points with $|\Delta| >$ `edgeLabelThreshold` are labelled.  Note that with only
8 subjects per stratum, many edges will have low and similar bootstrap strength —
a real-study comparison plot based on $n \approx 30+$ per stratum is far more
discriminating.

```{r plot-exposure-comparison, fig.width=8.5, fig.height=6}
pipeline_result$plotObjects$stratifiedModels$exposure$comparisonPlots[["Malaria Preexposed_vs_Malaria Naive"]]
```

### 11.3 Comparison table

The underlying data for the comparison plot are in the `comparisons` list.

```{r comparison-table}
comp <- pipeline_result$plotObjects$stratifiedModels$exposure$comparisons
comp_tbl <- comp[["Malaria Preexposed_vs_Malaria Naive"]]$comparisonTable

knitr::kable(
  head(comp_tbl[order(-comp_tbl$absoluteDelta), ], 10),
  digits = 3,
  row.names = FALSE,
  caption = "Top 10 edges ranked by |delta| — Malaria Preexposed vs. Naive"
)
```

Column definitions:

| Column | Meaning |
|--------|---------|
| `edge` | Undirected edge (alphabetical node order) |
| `firstStrength` | Raw bootstrap stability in Malaria Preexposed |
| `secondStrength` | Raw bootstrap stability in Malaria Naive |
| `firstSignedStrength` | `sign(r) × stability` in Malaria Preexposed |
| `secondSignedStrength` | `sign(r) × stability` in Malaria Naive |
| `delta` | `firstSignedStrength − secondSignedStrength` |
| `absoluteDelta` | `|delta|` (magnitude of difference) |
| `direction` | "Stronger in {group}" / "Similar" |
| `edgeType` | Edge category (`Exogenous to measurement`, `Within {groupName}`, `Cross-compartment`) |

### 11.4 Protection stratification: Protected vs. Not protected

Only **vaccinated** subjects with a **non-missing** protection outcome are eligible.
This filters to S01–S16 (16 subjects), partitioned by `Protection` (8 "Yes", 8 "No").

```{r plot-protection-yes, fig.width=9.5, fig.height=7}
pipeline_result$plotObjects$stratifiedModels$protection$levelPlots[["Yes"]]
```

```{r plot-protection-comparison, fig.width=8.5, fig.height=6}
pipeline_result$plotObjects$stratifiedModels$protection$comparisonPlots[["Yes_vs_No"]]
```

---

## 12. Interpreting results

### 12.1 Reading bootstrap strength

| Strength | Interpretation |
|----------|---------------|
| < 0.10 | Unreliable — likely noise |
| 0.10–0.25 | Weak evidence; below typical thresholds |
| 0.25–0.50 | Moderate evidence; worth investigating |
| ≥ 0.50 | Strong evidence of a conditional dependency |
| ≥ 0.80 | Very strong; edge present in the majority of bootstrap networks |

The recommended production threshold for the **averaged network** is **0.40**.  We
use 0.20 in this vignette only because the demo dataset is deliberately small.

### 12.2 Signed strength and the delta score

Raw bootstrap strength is non-negative.  Multiplying by $\text{sign}(r)$ gives
**signed strength**, which captures whether the dependency is positive or negative
(e.g. a consistent *inverse* co-variation between two cell types has $r < 0$ and
therefore negative signed strength despite potentially high bootstrap stability).

The **delta** $\Delta = s_1 - s_2$ quantifies differential connectivity:

- **Large positive $\Delta$**: edge is stronger/more positive in Group 1.
- **Large negative $\Delta$**: edge is stronger/more positive in Group 2.
- **$|\Delta|$ near zero**: edge is similar across groups, or absent in both.

The default labelling threshold (`edgeLabelThreshold`) is **0.20** — only edges
with $|\Delta| > 0.20$ receive labels in the comparison scatter plot.

### 12.3 Edge types

`classify_edge_type()` assigns each edge in the comparison table to one of the
following categories:

| Edge type | Example | Biological meaning |
|-----------|---------|-------------------|
| `Exogenous to measurement` | `GroupName → Liver_Pct_Kupffer_Cells` | Treatment group predicts cell fraction |
| `Within liverCells` / `Within pbmcCells` | `Liver_Pct_NK_Cells → Liver_Pct_CD8_T_Cells` | Co-variation within a compartment |
| `Cross-compartment` | `Liver_Pct_Kupffer_Cells → PBMC_Pct_CD8_T_Cells` | Hepatic state predicts systemic immunity |
| `Exogenous to exogenous` | Blacklisted; should not appear | — |

Cross-compartment edges are the most biologically interesting because they encode inferred
information flow between tissue compartments.

### 12.4 Important caveats

1. **Conditional independence ≠ causation.** A DAG edge $A \to B$ means only that
   $A$ is a direct conditional predictor of $B$ in the model.  Causal interpretation
   requires additional domain assumptions and experimental validation.
2. **Small samples inflate bootstrap instability.** With $n < 20$ per stratum, many
   edges will have stability < 0.30 regardless of the true underlying structure.
3. **Cross-compartment inference still depends on missing-data handling.** Each row now
  contains real paired Liver and PBMC measurements for the same subject, so MICE uses
  genuine within-subject cross-compartment information when filling assay dropouts.
  Any imputed values reflect assay failure, not a structurally absent tissue block.
4. **Discretization destroys some signal.** Three-level discretization loses
   continuous variation.  Edges between variables with complex non-linear relationships
   may be missed or artificially strengthened.

### 12.5 Recommended production settings

```{r production-settings, eval=FALSE}
# Production-grade overrides — replace the demo values above
production_overrides <- list(
  structureLearning = list(
    bootstrapReplicates     = 1000,   # Friedman et al. recommend R >= 200, ideally >= 500
    averageNetworkThreshold = 0.40    # Conservative; increases edge specificity
  ),
  imputation = list(
    numberOfImputations = 5,          # Standard MICE recommendation
    maximumIterations   = 20
  ),
  discretization = list(
    initialBreaks = 20,               # Default; adequate for n >= 30
    finalBreaks   = 3
  )
)
```

With $n \approx 50$+ per sub-cohort and 1 000 bootstrap replicates, a full pipeline
run typically completes in 5–15 minutes on a modern laptop.

---

## Session information

```{r session-info}
sessionInfo()
```
